---
title: "README"
author: "Ramzan Kamoto"
date: "2025-01-31"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, results='hide', echo=FALSE}

rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
suppressPackageStartupMessages(pacman::p_load(tidyverse))
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```



# Risk Parity Portfolio construction in SA: README

## Purpose 

This is the host platform for my Fin metrics essay. Initially I wanted to work on active share and fund flows but i was scrambling for ideas so I switched to something where code was available. The essay was written using rticles elsevier template. This is because after many attempts, Texevier would just not work for me. I try to mimic the texevier format as far as possible. I trust this is okay. 

In this essay I attempt to construct a risk parity portfolio and compare it with a tangency portfolio as done by Palomar and T.Souza. Essentially I  undertake a conceptual replication in that I take their ideas, code and packages but apply it to SA data. The data used was the J200 datset from fmxdat. In the Palomar vignette, they use FAANG companies to illustrate risk parity construction. Similarly, I take the top 10 tickers on the last date and use those for the rest of my analysis. 

Some things I was able to do well and others I struggled. First, I was able to get nice plots showing the weight and risk contribution under risk parity portfolio conditions. I was also able to compare risk parity and tangency portfolio weights side by side. constructing the tangency portfolio proved difficult. I initially tried to follow the Palomar vignette but my portfolio failed to optimize. Then, I opted for CVXR which worked but then I struggled to get the rolling weights stacked bar chart. In the end I abandoned the tangency portfolio at this stage. In the essay, I discuss based on what I have seen in the literature but not with a plot of my own. 


## Code issue 

Very close to the submission time as I was tying my work together my function to load data started to produce an error (Error in `>.default`(date, start_date) : 
  comparison (>) is possible only for atomic and list types)

I tried stack and google, some suggestions were that it is because of lubridates date function being confused but I am not sure. It could just be my machine. Please do review the code and the plots that are unfortunately missing from my essay.  
  
## Initializing


```{r, results='hide', echo=TRUE}

suppressPackageStartupMessages({
  pacman::p_load(tidyverse, RiskPortfolios, fitHeavyTail)
  library(lubridate)
  library(tbl2xts)
  library(riskParityPortfolio)
  library(portfolioBacktest)
  library(xts)
  library(fPortfolio)
  library(PerformanceAnalytics)
  library(fmxdat)
})

```
# Load Data 

```{r}
# Start_date <- ymd(20100103)
# J200 <- load_data(Start_date)
# return_mat <- prepare_return_matrix(J200)
# return_mat <- impute_missing_returns(return_mat, method = "Drawn_Distribution_Collective") %>% tbl_xts()
# 

```


# Table of statistics 

```{r}
# generate_table1(return_mat)

```

# Risk Contribution plot 

```{r}
# plot_risk_parity(return_mat)

```

# Parity vs Tangency weights 

```{r}

# plot_portfolio_weights(return_mat)

```

# Rolling risk parity 

```{r}
# plot_rolling_risk_parity(return_mat)

```

# Performance chart for risk parity index 

```{r}
# generate_performance_summary(return_mat)

```

